---
title: "Simple Normal Regression"
format: html
editor: visual
---

## [Simple Normal Regression](https://www.bayesrulesbook.com/chapter-9#:~:text=Chapter%209-,Simple%20Normal%20Regression,-Welcome%20to%20Unit)

```{r}
library(bayesrules)
library(tidyverse)
library(rstan)
library(rstanarm)
library(bayesplot)
library(tidybayes)
library(janitor)
library(broom.mixed)
library(ggpubr)
library(ggplot2)
options(scipen = 99)
```

### Case Study - Capital Bikeshare

We want to understand the relationship between temperature and ridership

```{r}
data(bikes)
```

💻 Do some exploratory data viz to tell me about the relationship between `temp_feel` and `rides`. Also make a guess for the number of rides based on `temp_feel` being 75 degrees

```{r}
ggplot(bikes, aes(x = temp_feel, y = rides)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "loess", color = "darkred", se = TRUE) +
  labs(
    title = "Relationship Between Feels-Like Temperature and Bike Rides",
    x = "Feels-Like Temperature (°F)",
    y = "Number of Rides"
  ) +
  theme_minimal()

```

```{r}
cor(bikes$temp_feel, bikes$rides, use = "complete.obs")
```

So the frequentest statistical methods allow us to calculate a correlation coefficient to measure the strength/direction of the relationship and a p value to measure the significance

```{r}
ggplot(data = bikes, aes(x = temp_feel, y = rides)) +
  geom_point() +
  geom_smooth() +
  stat_cor()
```

```{r}
fit <- lm(rides ~ temp_feel, bikes)
fit
```

This linear regression model tells us that for each 1 degree increase in temperature there is an expected increase of 81 rides

```{r}
summary(fit)$coefficients
```

```{r}
confint(fit, "temp_feel", level = 0.95) 
```

“Using `lm()`, the 95% confidence interval for the slope of `temp_feel` is **\[71.82, 91.94\]** rides per °F. If we repeated this study many times and computed intervals the same way, **95% of those intervals** would contain the true slope.”

### How do we approach this example with Bayesian statistics?

First, we'll need priors:

1.  On an *average* temperature day, say 65 or 70 degrees for D.C., there are typically around 5000 riders, though this average could be somewhere between 3000 and 7000.

```{r}
plot_normal(mean = 5000, sd = 1000) + 
  labs(x = "beta_0c", y = "pdf") +
  geom_vline(xintercept = 3000) +
  geom_vline(xintercept = 7000) 
```

1.  For every one degree increase in temperature, ridership typically increases by 100 rides, though this average increase could be as low as 20 or as high as 180.

```{r}
plot_normal(mean = 100, sd = 40) + 
  labs(x = "beta_1", y = "pdf") +
  geom_vline(xintercept = 20) +
  geom_vline(xintercept = 180) 
```

1.  At any given temperature, daily ridership will tend to vary with a moderate standard deviation of 1250 rides.

    -   For an Exponential, the average equals 1 / rate. So set rate = 1 / 1250 = 0.0008.

```{r}
plot_gamma(shape = 1, rate = 0.0008) + 
  labs(x = "sigma", y = "pdf")
```

### Simulation via rstanarm

```{r}
bike_model <- stan_glm(rides ~ temp_feel, data = bikes,
                       family = gaussian,
                       prior_intercept = normal(5000, 1000),
                       prior = normal(100, 40), 
                       prior_aux = exponential(0.0008),
                       chains = 4, iter = 5000*2, seed = 84735)
```

```{r}
bike_model
```

-   **Model & data**: You fit a straight-line model of **rides** vs **temp_feel** using **500 days**.

-   **Slope (temp_feel = 82.2, MAD_SD ≈ 5.1)**\

-   On average, **each +1°F** is associated with about **+82 more rides**. The small MAD_SD means the slope is estimated pretty precisely and is almost certainly positive.

-   **Intercept ((Intercept) = −2195.3, MAD_SD ≈ 354)**\
    This is where the fitted line crosses the y-axis at **0°F**. It’s not meant to be literal (negative rides); it just anchors the line. If you center temperature (e.g., subtract the average temp), the intercept would become “typical rides at an average-temp day,” which is more interpretable.

-   **Residual spread (sigma = 1282.5, MAD_SD ≈ 40)** Even **at the same temperature**, daily rides bounce around the line by about **±1,280 rides** on average. This real-world volatility dominates single-day predictions.

“From `stan_glm`, the slope’s 95% **credible interval** is **about \[72.2, 92.2\]** rides per °F (using Median = 82.2 and MAD_SD = 5.1). **Given the model, data, and priors, there’s a 95% probability** that the true slope lies in this range.”

## The big idea

-   **Confidence interval (CI)** — *frequentist*\
    A 95% CI is a **procedure** that, if you repeated the whole data-collection process forever, would produce intervals that **cover the fixed, unknown parameter 95% of the time**. After you’ve seen *this* data, the parameter is not random; the random thing was the interval-making procedure.\

    **You may NOT say:** “there’s a 95% chance the true value is in this specific CI.”

-   **Credible interval (CrI)** — *Bayesian*\

    A 95% CrI is the set of parameter values that have **95% posterior probability** *given your data and your prior*. After you’ve seen the data, it’s valid to say:\

    **You may say:** “there’s a 95% probability the true value lies in this interval (given model + prior).”

## Posterior prediction

Suppose a weather report indicates that tomorrow will be a 75-degree day in D.C. What’s your posterior guess of the number of riders that Capital Bikeshare should anticipate?

```{r}
# Simulate a set of predictions
set.seed(84735)
prediction <- posterior_predict(bike_model, newdata = data.frame(temp_feel = 75))
```

```{r}
posterior_interval(prediction, prob = 0.95)
```

```{r}
# Plot the approximate predictive model
mcmc_dens(prediction) + 
  xlab("predicted ridership on a 75 degree day") +
  geom_vline(xintercept = 1492, color = "coral") +
  geom_vline(xintercept = 6502, color = "coral") +
  theme_minimal()
```

**Interpretation:** Given your model + priors + the 500 days of data, there’s a **95% probability** that on a 75 degree day ridership will fall between **1,492 and 6,503** rides

How do we feel about the uncertainty in this prediction?

```{r}
# Bayesian crochet tension demo (yarn weight as predictor)
# Required packages:
# install.packages(c("bayesrules","tidyverse","rstanarm","bayesplot")) 
# or use brms if you prefer.

library(bayesrules)   # convenience helpers for teaching Bayesian workflows
library(tidyverse)
library(rstanarm)     # or brms
library(bayesplot)

set.seed(2025)

# -----------------------------
# 1) Simulate theoretical data
# -----------------------------
# We'll measure yarn weight as grams-per-meter (g_per_m)
n <- 60
yarn_gpm <- runif(n, 50, 400)   # 50 g/m (very thin) to 400 g/m (very thick)
# Assume thicker yarn slightly reduces tension variability:
beta0_true <- 2.0    # baseline tension_sd in mm
beta1_true <- -0.03  # per 1 g/m; so 10 g/m -> -0.3 mm
sigma_true <- 0.45

tension_sd <- beta0_true + beta1_true * yarn_gpm + rnorm(n, 0, sigma_true)

crochet <- tibble(yarn_gpm = yarn_gpm,
                  tension_sd = tension_sd)

# Quick plot of simulated data
ggplot(crochet, aes(x = yarn_gpm, y = tension_sd)) +
  geom_point() +
  labs(x = "Yarn weight (g/m)", y = "Stitch length SD (mm)",
       title = "Simulated tension data (theoretical)")

# -----------------------------
# 2) Priors
# -----------------------------
# weakly informative priors centered near 0
prior_intercept <- normal(location = 0, scale = 2)  # intercept
prior_coef      <- normal(location = 0, scale = 1)  # slope
prior_sigma     <- exponential(rate = 1)            # error

# -----------------------------
# 3) Fit Bayesian linear regression (rstanarm)
# -----------------------------
fit <- stan_glm(
  tension_sd ~ yarn_gpm,
  data = crochet,
  family = gaussian(),
  prior_intercept = prior_intercept,
  prior = prior_coef,
  prior_aux = prior_sigma,
  chains = 4, iter = 4000, refresh = 0, seed = 2025
)

print(fit)
posterior_samples <- as.matrix(fit)

# -----------------------------
# 4) Posterior summaries & plots
# -----------------------------
# Posterior summary
posterior_summary <- summary(fit, digits = 3)
posterior_summary

# Plot posterior for slope
mcmc_areas(as.matrix(fit)[, c("(Intercept)", "yarn_gpm")],
           prob = 0.95) +
  ggtitle("Posterior distributions (Intercept and yarn_gpm)")

# Posterior predictive regression line + credible interval
new_x <- tibble(yarn_gpm = seq(min(crochet$yarn_gpm), max(crochet$yarn_gpm), length.out = 100))
preds <- posterior_predict(fit, newdata = new_x)
pred_mean <- apply(preds, 2, mean)
pred_lower <- apply(preds, 2, quantile, probs = 0.025)
pred_upper <- apply(preds, 2, quantile, probs = 0.975)

pred_df <- bind_cols(new_x,
                     tibble(pred_mean = pred_mean,
                            pred_lower = pred_lower,
                            pred_upper = pred_upper))

ggplot(crochet, aes(y = tension_sd, x = yarn_gpm)) +
  geom_point(alpha = 0.6) +
  geom_line(data = pred_df, aes(x = yarn_gpm, y = pred_mean), size = 1) +
  geom_ribbon(data = pred_df, aes(x = yarn_gpm, ymin = pred_lower, ymax = pred_upper), alpha = 0.15) +
  labs(x = "Yarn weight (g/m)", y = "Stitch length SD (mm)",
       title = "Posterior predictive regression and 95% credible band")

# -----------------------------
# 5) Interpretive summary (print)
# -----------------------------
beta1_post_mean <- mean(posterior_samples[, "yarn_gpm"])
beta1_post_ci   <- quantile(posterior_samples[, "yarn_gpm"], probs = c(0.025, 0.975))
p_beta1_neg     <- mean(posterior_samples[, "yarn_gpm"] < 0)

cat("Posterior mean for beta1 (yarn_gpm):", round(beta1_post_mean, 3), "\n")
cat("95% credible interval for beta1:", round(beta1_post_ci[1],3), "to", round(beta1_post_ci[2],3), "\n")
cat("Posterior probability beta1 < 0:", round(p_beta1_neg, 3), "\n")

```

```{r}
library(bayesrules)
library(tidyverse)
library(rstanarm)

model <- stan_glm(
  tension_sd ~ yarn_gpm,
  data = crochet,
  prior_intercept = normal(0, 2),
  prior = normal(0, 1),
  prior_aux = exponential(1)
)
summary(model)

```
